---
Title of code: "3. Male Elephant Character Durability - Repeatability Models"
Author of code: "Jodie L. Berezin"
Date last updated: "May 30th, 2024"

Manuscript: "Dynamics of male African elephant character durability across time, age, and social context"
Authors of manuscript: "Caitlin E. O'Connell-Rodwell, Jodie L. Berezin, Colleen Kinzley, Patrick T. Freeman, Monica N. Sandri, Dustin Kieschnick, Mariana Abarca, Virginia Hayssen"
---

*About this code:*
The script below contains the repeatability models used to test for consistency in male elephant behavior across time and social context.

The script begins after data wrangling. Data is currently not provided due to being part of an on-going study.

Any questions about the data or code can be directed to the corresponding authors:
CEO-R, ceoconnell[at]stanford.edu
JLB, jlberezin[at]gmail.com

*Table of Contents*
1. Checking distributions for each behavior
2. Removing individuals with only one rate for each behavior
3. Modeling
  - Model prep notes
  - Models for each behavior


# 1. Checking distributions for each behavior

*All need log behav rates*
```{r}
# affiliation - 408 obvs
affil.mushara.behav.rates3 <- mushara.behav.rates3 %>%
  filter(BehavCat == "affiliation")

hist(affil.mushara.behav.rates3$BehavRate) # right skew
hist(affil.mushara.behav.rates3$LogBehavRate) # normal - use log


# aggression - 429 obvs
aggr.mushara.behav.rates3 <- mushara.behav.rates3 %>%
  filter(BehavCat == "aggression")

hist(aggr.mushara.behav.rates3$BehavRate) # right skew
hist(aggr.mushara.behav.rates3$LogBehavRate) # normal - use log


# dominance 314 obvs
dom.mushara.behav.rates3 <- mushara.behav.rates3 %>%
  filter(BehavCat == "dominance")

hist(dom.mushara.behav.rates3$BehavRate) # right skew
hist(dom.mushara.behav.rates3$LogBehavRate) # normal - use log


# escalated aggression - 58 obvs
escaggr.mushara.behav.rates3 <- mushara.behav.rates3 %>%
  filter(BehavCat == "esc.aggr")

hist(escaggr.mushara.behav.rates3$BehavRate) # right skew
hist(escaggr.mushara.behav.rates3$LogBehavRate) # still not super normal but I think it's okay


# play - 160 obvs
play.mushara.behav.rates3 <- mushara.behav.rates3 %>%
  filter(BehavCat == "play")

hist(play.mushara.behav.rates3$BehavRate) # right skew
hist(play.mushara.behav.rates3$LogBehavRate) # normal - use log


# retreat - 116 obvs
retreat.mushara.behav.rates3 <- mushara.behav.rates3 %>%
  filter(BehavCat == "retreat")

hist(retreat.mushara.behav.rates3$BehavRate) # right skew
hist(retreat.mushara.behav.rates3$LogBehavRate) # normal - use log


# SD anxious - 414 obvs
sda.mushara.behav.rates3 <- mushara.behav.rates3 %>%
  filter(BehavCat == "sd.anxious")

hist(sda.mushara.behav.rates3$BehavRate) # right skew
hist(sda.mushara.behav.rates3$LogBehavRate) # normal - use log


# SD comfort - 160 obvs
sdc.mushara.behav.rates3 <- mushara.behav.rates3 %>%
  filter(BehavCat == "sd.comfort")

hist(sdc.mushara.behav.rates3$BehavRate) # right skew
hist(sdc.mushara.behav.rates3$LogBehavRate) # mostly normal - use log


# social contentment - 364 obvs
soc.mushara.behav.rates3 <- mushara.behav.rates3 %>%
  filter(BehavCat == "social.content")

hist(soc.mushara.behav.rates3$BehavRate) # right skew
hist(soc.mushara.behav.rates3$LogBehavRate) # normal - use log


# vigilance - 502 obvs
vig.mushara.behav.rates3 <- mushara.behav.rates3 %>%
  filter(BehavCat == "vigilance")

hist(vig.mushara.behav.rates3$BehavRate) # right skew
hist(vig.mushara.behav.rates3$LogBehavRate) # normal - use log
```


# 2. Removing individuals with only one rate for each behavior

Summarized to see who is being removed, then they were removed in a second step.
```{r}
# affiliation - no change, still 408 obvs
affil.rates.indiv.count <- affil.mushara.behav.rates3 %>%
  group_by(ElephantName) %>%
  summarise(n = n()) %>%
  mutate(keep = ifelse(n < 2, "remove", "keep"))
affil.rates.indiv.count


# aggression - no change, still 429 obvs
aggr.rates.indiv.count <- aggr.mushara.behav.rates3 %>%
  group_by(ElephantName) %>%
  summarise(n = n()) %>%
  mutate(keep = ifelse(n < 2, "remove", "keep"))
aggr.rates.indiv.count


# dominance - removed four individuals, now 310 rates
dom.rates.indiv.count <- dom.mushara.behav.rates3 %>%
  group_by(ElephantName) %>%
  summarise(n = n()) %>%
  mutate(keep = ifelse(n < 2, "remove", "keep"))
dom.rates.indiv.count # four to remove

dom.remove.indiv <- c("144", "64", "146", "143")
dom.mushara.behav.rates3.2 <- dom.mushara.behav.rates3 %>%
  filter(!ElephantName %in% dom.remove.indiv) # only 4 removed - great!


# escalated aggression - removed six individuals, now 52 obvs
escag.rates.indiv.count <- escaggr.mushara.behav.rates3 %>%
  group_by(ElephantName) %>%
  summarise(n = n()) %>%
  mutate(keep = ifelse(n < 2, "remove", "keep"))
escag.rates.indiv.count # six to remove

escaggr.remove.indiv <- c("62", "48", "101", "37", "111", "25")
escaggr.mushara.behav.rates3.2 <- escaggr.mushara.behav.rates3 %>%
  filter(!ElephantName %in% escaggr.remove.indiv)


# play - removed six individuals, now 154 rates
play.rates.indiv.count <- play.mushara.behav.rates3 %>%
  group_by(ElephantName) %>%
  summarise(n = n()) %>%
  mutate(keep = ifelse(n < 2, "remove", "keep"))
play.rates.indiv.count # 6 to remove

play.remove.indiv <- c("144", "140", "77", "92", "143", "70")
play.mushara.behav.rates3.2 <- play.mushara.behav.rates3 %>%
  filter(!ElephantName %in% play.remove.indiv)


# retreat -removed 10 individuals, now 106 obvs
retreat.rates.indiv.count <- retreat.mushara.behav.rates3 %>%
  group_by(ElephantName) %>%
  summarise(n = n()) %>%
  mutate(keep = ifelse(n < 2, "remove", "keep"))
retreat.rates.indiv.count # 10 to remove

retreat.remove.indiv <- c("144", "112", "65", "30", "77", "93", "69", "146", "143", "96")
retreat.mushara.behav.rates3.2 <- retreat.mushara.behav.rates3 %>%
  filter(!ElephantName %in% retreat.remove.indiv)


# SD anxious - no change, still with 414 obvs
sda.rates.indiv.count <- sda.mushara.behav.rates3 %>%
  group_by(ElephantName) %>%
  summarise(n = n()) %>%
  mutate(keep = ifelse(n < 2, "remove", "keep"))
sda.rates.indiv.count # none to remove


# SD comfort - six individuals to remove, now 154 obvs
sdc.rates.indiv.count <- sdc.mushara.behav.rates3 %>%
  group_by(ElephantName) %>%
  summarise(n = n()) %>%
  mutate(keep = ifelse(n < 2, "remove", "keep"))
sdc.rates.indiv.count # 6 to remove

sdc.remove.indiv <- c("101", "69", "92", "6", "96", "70")
sdc.mushara.behav.rates3.2 <- sdc.mushara.behav.rates3 %>%
  filter(!ElephantName %in% sdc.remove.indiv)


# social content - two individuals to remove, now 362 obvs
soc.rates.indiv.count <- soc.mushara.behav.rates3 %>%
  group_by(ElephantName) %>%
  summarise(n = n()) %>%
  mutate(keep = ifelse(n < 2, "remove", "keep"))
soc.rates.indiv.count # 2 to remove

soc.remove.indiv <- c("92", "83")
soc.mushara.behav.rates3.2 <- soc.mushara.behav.rates3 %>%
  filter(!ElephantName %in% soc.remove.indiv)


# vigilance - no change, still 502 obvs
vig.rates.indiv.count <- vig.mushara.behav.rates3 %>%
  group_by(ElephantName) %>%
  summarise(n = n()) %>%
  mutate(keep = ifelse(n < 2, "remove", "keep"))
vig.rates.indiv.count # none to remove
```


# 3. Modeling

Notes:
I used the Our Coding Club tutorial describing linear mixed models and how to specify random effects: https://ourcodingclub.github.io/tutorials/mixed-models/#types

The models will have the (log) behavior rate as the response variable and three random effects to test repeatability. Here is the rationale for setting up each models random effects:
- Individual elephants are observed in multiple years and multiple events. Since no elephant is observed ONLY within one of these samples, then elephant is a cross effect with year and event (specified separately) from the other random effects. +(1|ElephantName)
- Our years and events are tied in with each other as well: "Our site variable is a three-level factor, with sites called a, b and c. The nesting of the site within the mountain range is implicit - our sites are meaningless without being assigned to specific mountain ranges, i.e. there is nothing linking site b of the Bavarian mountain range with site b of the Central mountain range." This is true of our events within the years - there is nothing linking event 5 of 2007 and event 5 of 2009. This is an *implicit* effect and we can make *explicit* by specifying a new variable in our data (done in previous steps, YEvent = Year:Event).


## Affiliation
n = 408 rates, individuals = 34
```{r}
affil.rptr <- rpt(LogBehavRate ~ (1|ElephantName) + (1|Year) + (1|YEvent), 
                                     grname = c("ElephantName", "Year", "YEvent"), data = affil.mushara.behav.rates3, 
                  datatype = "Gaussian", nboot = 1000, npermut = 0)

print(affil.rptr) 

plot(affil.rptr, grname = "ElephantName", type = "boot", cex.main = 0.8)
plot(affil.rptr, grname = "Year", type = "boot", cex.main = 0.8)
plot(affil.rptr, grname = "YEvent", type = "boot", cex.main = 0.8)

## checking the model
plot(residuals(affil.rptr$mod)) # linearity and equality of variance - pretty random
ggqqplot(residuals(affil.rptr$mod)) # normality of residuals - looks good
hist(residuals(affil.rptr$mod)) # normal
```

## Aggression
n = 429 rates, individuals = 34
```{r}
aggr.rptr <- rpt(LogBehavRate ~ (1|ElephantName) + (1|Year) + (1|YEvent), 
                                     grname = c("ElephantName", "Year", "YEvent"), data = aggr.mushara.behav.rates3, 
                  datatype = "Gaussian", nboot = 1000, npermut = 0)

print(aggr.rptr) 

plot(aggr.rptr, grname = "ElephantName", type = "boot", cex.main = 0.8)
plot(aggr.rptr, grname = "Year", type = "boot", cex.main = 0.8)
plot(aggr.rptr, grname = "YEvent", type = "boot", cex.main = 0.8)

## checking the model
plot(residuals(aggr.rptr$mod)) # linearity and equality of variance - pretty random
ggqqplot(residuals(aggr.rptr$mod)) # normality of residuals - looks good
hist(residuals(aggr.rptr$mod)) # normal
```


## Dominance
n = 310 rates, individuals = 30
```{r}
dom.rptr <- rpt(LogBehavRate ~ (1|ElephantName) + (1|Year) + (1|YEvent), 
                                     grname = c("ElephantName", "Year", "YEvent"), data = dom.mushara.behav.rates3.2, 
                  datatype = "Gaussian", nboot = 1000, npermut = 0)

print(dom.rptr) 

plot(dom.rptr, grname = "ElephantName", type = "boot", cex.main = 0.8)
plot(dom.rptr, grname = "Year", type = "boot", cex.main = 0.8)
plot(dom.rptr, grname = "YEvent", type = "boot", cex.main = 0.8)

## checking the model
plot(residuals(dom.rptr$mod)) # linearity and equality of variance - pretty random
ggqqplot(residuals(dom.rptr$mod)) # normality of residuals - looks good
hist(residuals(dom.rptr$mod)) # normal
```


## Escalated Aggression
n = 52 rates, individuals = 16
```{r}
escag.rptr <- rpt(LogBehavRate ~ (1|ElephantName) + (1|Year) + (1|YEvent), 
                                     grname = c("ElephantName", "Year", "YEvent"), data = escaggr.mushara.behav.rates3.2, 
                  datatype = "Gaussian", nboot = 1000, npermut = 0)

print(escag.rptr) 

plot(escag.rptr, grname = "ElephantName", type = "boot", cex.main = 0.8)
plot(escag.rptr, grname = "Year", type = "boot", cex.main = 0.8)
plot(escag.rptr, grname = "YEvent", type = "boot", cex.main = 0.8)

## checking the model
plot(residuals(escag.rptr$mod)) # linearity and equality of variance - pretty random
ggqqplot(residuals(escag.rptr$mod)) # normality of residuals - looks good
hist(residuals(escag.rptr$mod)) # normal
```


## Play
n = 154 rates, individuals = 26
```{r}
play.rptr <- rpt(LogBehavRate ~ (1|ElephantName) + (1|Year) + (1|YEvent), 
                                     grname = c("ElephantName", "Year", "YEvent"), data = play.mushara.behav.rates3.2, 
                  datatype = "Gaussian", nboot = 1000, npermut = 0)

print(play.rptr) 

plot(play.rptr, grname = "ElephantName", type = "boot", cex.main = 0.8)
plot(play.rptr, grname = "Year", type = "boot", cex.main = 0.8)
plot(play.rptr, grname = "YEvent", type = "boot", cex.main = 0.8)

## checking the model
plot(residuals(play.rptr$mod)) # linearity and equality of variance - pretty random
ggqqplot(residuals(play.rptr$mod)) # normality of residuals - looks good
hist(residuals(play.rptr$mod)) # normal
```


## Retreat
n = 106 rates, individuals = 21
```{r}
retreat.rptr <- rpt(LogBehavRate ~ (1|ElephantName) + (1|Year) + (1|YEvent), 
                                     grname = c("ElephantName", "Year", "YEvent"), data = retreat.mushara.behav.rates3.2, 
                  datatype = "Gaussian", nboot = 1000, npermut = 0)

print(retreat.rptr) 

plot(retreat.rptr, grname = "ElephantName", type = "boot", cex.main = 0.8)
plot(retreat.rptr, grname = "Year", type = "boot", cex.main = 0.8)
plot(retreat.rptr, grname = "YEvent", type = "boot", cex.main = 0.8)

## checking the model
plot(residuals(retreat.rptr$mod)) # linearity and equality of variance - pretty random
ggqqplot(residuals(retreat.rptr$mod)) # normality of residuals - looks good
hist(residuals(retreat.rptr$mod)) # normal
```


## Self-directed anxious
n = 414 rates, individuals = 34
```{r}
sda.rptr <- rpt(LogBehavRate ~ (1|ElephantName) + (1|Year) + (1|YEvent), 
                                     grname = c("ElephantName", "Year", "YEvent"), data = sda.mushara.behav.rates3, 
                  datatype = "Gaussian", nboot = 1000, npermut = 0)

print(sda.rptr) 

plot(sda.rptr, grname = "ElephantName", type = "boot", cex.main = 0.8)
plot(sda.rptr, grname = "Year", type = "boot", cex.main = 0.8)
plot(sda.rptr, grname = "YEvent", type = "boot", cex.main = 0.8)

## checking the model
plot(residuals(sda.rptr$mod)) # linearity and equality of variance - pretty random
ggqqplot(residuals(sda.rptr$mod)) # normality of residuals - looks good
hist(residuals(sda.rptr$mod)) # normal
```


## Self-directed comfort
n = 154 rates, individuals = 24
```{r}
sdc.rptr <- rpt(LogBehavRate ~ (1|ElephantName) + (1|Year) + (1|YEvent), 
                                     grname = c("ElephantName", "Year", "YEvent"), data = sdc.mushara.behav.rates3.2, 
                  datatype = "Gaussian", nboot = 1000, npermut = 0)

print(sdc.rptr) 

plot(sdc.rptr, grname = "ElephantName", type = "boot", cex.main = 0.8)
plot(sdc.rptr, grname = "Year", type = "boot", cex.main = 0.8)
plot(sdc.rptr, grname = "YEvent", type = "boot", cex.main = 0.8)

## checking the model
plot(residuals(sdc.rptr$mod)) # linearity and equality of variance - pretty random
ggqqplot(residuals(sdc.rptr$mod)) # normality of residuals - looks good
hist(residuals(sdc.rptr$mod)) # normal
```


## Social contentment
n = 362 rates, individuals = 32
```{r}
soc.rptr <- rpt(LogBehavRate ~ (1|ElephantName) + (1|Year) + (1|YEvent), 
                                     grname = c("ElephantName", "Year", "YEvent"), data = soc.mushara.behav.rates3.2, 
                  datatype = "Gaussian", nboot = 1000, npermut = 0)

print(soc.rptr)

plot(soc.rptr, grname = "ElephantName", type = "boot", cex.main = 0.8)
plot(soc.rptr, grname = "Year", type = "boot", cex.main = 0.8)
plot(soc.rptr, grname = "YEvent", type = "boot", cex.main = 0.8)

## checking the model
plot(residuals(soc.rptr$mod)) # linearity and equality of variance - pretty random
ggqqplot(residuals(soc.rptr$mod)) # normality of residuals - looks good
hist(residuals(soc.rptr$mod)) # normal
```


## Vigilance
n = 502 rates, individuals = 34
```{r}
vig.rptr <- rpt(LogBehavRate ~ (1|ElephantName) + (1|Year) + (1|YEvent), 
                                     grname = c("ElephantName", "Year", "YEvent"), data = vig.mushara.behav.rates3, 
                  datatype = "Gaussian", nboot = 1000, npermut = 0)

print(vig.rptr)

plot(vig.rptr, grname = "ElephantName", type = "boot", cex.main = 0.8)
plot(vig.rptr, grname = "Year", type = "boot", cex.main = 0.8)
plot(vig.rptr, grname = "YEvent", type = "boot", cex.main = 0.8)

## checking the model
plot(residuals(vig.rptr$mod)) # linearity and equality of variance - pretty random
ggqqplot(residuals(vig.rptr$mod)) # normality of residuals - looks good
hist(residuals(vig.rptr$mod)) # normal
```


