---
Title of code: "4. Male Elephant Character Durability - Linear Mixed-effect Models"
Author of code: "Jodie L. Berezin"
Date last updated: "May 31st, 2024"

Manuscript: "Dynamics of male African elephant character durability across time, age, and social context"
Authors of manuscript: "Caitlin E. O'Connell-Rodwell, Jodie L. Berezin, Colleen Kinzley, Patrick T. Freeman, Monica N. Sandri, Dustin Kieschnick, Mariana Abarca, Virginia Hayssen"
---

*About this code:*
The script below contains the linear mixed-effect models used to assess the impact that the presence of 1) musth males, 2) keystone male, and 3) youth males have on adult male (age classes 3Q, full, elder) behavior. The eight behaviors that were significantly repeatable-by-event were modeled (affiliation, aggression, dominance, play, retreat, self-directed anxious, social contentment, and vigilance).

The script begins after data wrangling. Data is currently not provided due to being part of an on-going study.

Any questions about the data or code can be directed to the corresponding authors:
CEO-R, ceoconnell[at]stanford.edu
JLB, jlberezin[at]gmail.com

*Table of Contents*
1. Checking distributions for each behavior
2. Modeling
3. Making Figure 2

~~~~~~~~~~~~~~~~~

# 1. Checking distributions for each behavior

Start by selecting the behavior categories, calculating the number of elephants and the number of behavior rates and then checking distributions (the natural log was calculated in a previous step in anticipation of skew).

## Affiliation
```{r}
affil.behav.rates <- bulls.behav.rates6 %>%
  filter(BehavCat == "affiliation") %>% # 265 rates
  droplevels()

unique(affil.behav.rates$ElephantName) # 23 individuals

hist(affil.behav.rates$BehavRate) # right skew
hist(affil.behav.rates$LogBehavRate) # normal!
```


## Aggression
```{r}
aggres.behav.rates <- bulls.behav.rates6 %>%
  filter(BehavCat == "aggression") %>% # 318 rates
  droplevels()

unique(aggres.behav.rates$ElephantName) # 23 individuals

hist(aggres.behav.rates$BehavRate) # right skew
hist(aggres.behav.rates$LogBehavRate) # normal!
```


## Dominance
```{r}
dom.behav.rates <- bulls.behav.rates6 %>%
  filter(BehavCat == "dominance") %>% # 216 rates
  droplevels()

unique(dom.behav.rates$ElephantName) # 23 individuals

hist(dom.behav.rates$BehavRate) # right skew
hist(dom.behav.rates$LogBehavRate) # normal!
```


## Play
```{r}
play.behav.rates <- bulls.behav.rates6 %>%
  filter(BehavCat == "play") %>% # 87 rates
  droplevels()

unique(play.behav.rates$ElephantName) # 20 individuals

hist(play.behav.rates$BehavRate) # right skew
hist(play.behav.rates$LogBehavRate, breaks = 5) # mostly normal - check residuals when modeling
```


## Retreat
```{r}
retreat.behav.rates <- bulls.behav.rates6 %>%
  filter(BehavCat == "retreat") %>% # 75 rates
  droplevels()

unique(retreat.behav.rates$ElephantName) # 21 individuals

hist(retreat.behav.rates$BehavRate) # right skew
hist(retreat.behav.rates$LogBehavRate) # normal!
```


## Self-directed anxious
```{r}
sda.behav.rates <- bulls.behav.rates6 %>%
  filter(BehavCat == "sd.anxious") %>% # 293 rates
  droplevels()

unique(sda.behav.rates$ElephantName) # 23 individuals

hist(sda.behav.rates$BehavRate) # right skew
hist(sda.behav.rates$LogBehavRate) # normal!
```


## Social contentment
```{r}
soc.behav.rates <- bulls.behav.rates6 %>%
  filter(BehavCat == "social.content") %>% # 261 rates
  droplevels()

unique(soc.behav.rates$ElephantName) # 23 individuals

hist(soc.behav.rates$BehavRate) # right skew
hist(soc.behav.rates$LogBehavRate) # normal!
```


## Vigilance
```{r}
vig.behav.rates <- bulls.behav.rates6%>%
  filter(BehavCat == "vigilance") %>% # 383 rates
  droplevels()

unique(vig.behav.rates$ElephantName) # 23 individuals

hist(vig.behav.rates$BehavRate) # right skew
hist(vig.behav.rates$LogBehavRate) # normal!
```


# 2. Modeling

We have to keep in mind the structure of the data - we have individuals that appear in many years, events, and intervals - nested within each other and partially crossed (see Methods section and Repeatability Models script for more details).

## Affiliation
```{r}
affil.lmer <- lmerTest::lmer(LogBehavRate ~ SocialContext + (1|Year/EventInterval) + (1|ElephantName), data = affil.behav.rates)

summary(affil.lmer)

# residuals
plot(affil.lmer)
qqnorm(resid(affil.lmer))
qqline(resid(affil.lmer))

# residuals
residuals.affil.lmer <- simulateResiduals(fittedModel = affil.lmer)
plot(residuals.affil.lmer) # looks good!!!

testDispersion(residuals.affil.lmer) # is good
testOutliers(residuals.affil.lmer, type = "bootstrap") # one possible outlier but n.s.
testZeroInflation(residuals.affil.lmer) # looks good!

affil.lmer.emm <- ggpredict(affil.lmer, terms = c("SocialContext")) # to be used for creating Figure 2
```



## Aggression
```{r}
aggres.lmer <- lmerTest::lmer(LogBehavRate ~ SocialContext + (1|Year/EventInterval) + (1|ElephantName), data = aggres.behav.rates)

summary(aggres.lmer)

# residuals
plot(aggres.lmer)
qqnorm(resid(aggres.lmer))
qqline(resid(aggres.lmer))

# residuals
residuals.aggres.lmer <- simulateResiduals(fittedModel = aggres.lmer)
plot(residuals.aggres.lmer) # looks good!!!

testDispersion(residuals.aggres.lmer) # is good
testOutliers(residuals.aggres.lmer, type = "bootstrap") # one significant outlier but the plot residuals says its fine - going to keep in the outlier
testZeroInflation(residuals.aggres.lmer) # looks good!

aggres.lmer.emm <- ggpredict(aggres.lmer, terms = c("SocialContext")) # to be used for creating Figure 2
```



## Dominance
```{r}
dom.lmer <- lmerTest::lmer(LogBehavRate ~ SocialContext + (1|Year/EventInterval) + (1|ElephantName), data = dom.behav.rates)

summary(dom.lmer)

# residuals
plot(dom.lmer)
qqnorm(resid(dom.lmer))
qqline(resid(dom.lmer))

# residuals
residuals.dom.lmer <- simulateResiduals(fittedModel = dom.lmer)
plot(residuals.dom.lmer) # looks good!!!

testDispersion(residuals.dom.lmer) # is good
testOutliers(residuals.dom.lmer, type = "bootstrap") # looks good
testZeroInflation(residuals.dom.lmer) # looks good!

dom.lmer.emm <- ggpredict(dom.lmer, terms = c("SocialContext")) # to be used for creating Figure 2
```



## Play
```{r}
play.lmer <- lmerTest::lmer(LogBehavRate ~ SocialContext + (1|Year/EventInterval) + (1|ElephantName), data = play.behav.rates)

summary(play.lmer)

# residuals
plot(play.lmer)
qqnorm(resid(play.lmer))
qqline(resid(play.lmer))

# residuals
residuals.play.lmer <- simulateResiduals(fittedModel = play.lmer)
plot(residuals.play.lmer) # looks good!!!

testDispersion(residuals.play.lmer) # is good
testOutliers(residuals.play.lmer, type = "bootstrap") # looks good
testZeroInflation(residuals.play.lmer) # looks good!

play.lmer.emm <- ggpredict(play.lmer, terms = c("SocialContext")) # to be used for creating Figure 2
```



## Retreat
```{r}
retreat.lmer <- lmerTest::lmer(LogBehavRate ~ SocialContext + (1|Year/EventInterval) + (1|ElephantName), data = retreat.behav.rates)

summary(retreat.lmer)

# residuals
plot(retreat.lmer)
qqnorm(resid(retreat.lmer))
qqline(resid(retreat.lmer))

# residuals
residuals.retreat.lmer <- simulateResiduals(fittedModel = retreat.lmer)
plot(residuals.retreat.lmer) # looks good!!!

testDispersion(residuals.retreat.lmer) # is good
testOutliers(residuals.retreat.lmer, type = "bootstrap") # looks good
testZeroInflation(residuals.retreat.lmer) # looks good!

retreat.lmer.emm <- ggpredict(retreat.lmer, terms = c("SocialContext")) # to be used for creating Figure 2
```



## Self-directed anxious
```{r}
sda.lmer <- lmerTest::lmer(LogBehavRate ~ SocialContext + (1|Year/EventInterval) + (1|ElephantName), data = sda.behav.rates)

summary(sda.lmer)

# residuals
plot(sda.lmer)
qqnorm(resid(sda.lmer))
qqline(resid(sda.lmer))

# residuals
residuals.sda.lmer <- simulateResiduals(fittedModel = sda.lmer)
plot(residuals.sda.lmer) # looks good!!!

testDispersion(residuals.sda.lmer) # is good
testOutliers(residuals.sda.lmer, type = "bootstrap") # looks good
testZeroInflation(residuals.sda.lmer) # looks good!

sda.lmer.emm <- ggpredict(sda.lmer, terms = c("SocialContext")) # to be used for creating Figure 2
```



## Social contentment
```{r}
soc.lmer <- lmerTest::lmer(LogBehavRate ~ SocialContext + (1|Year/EventInterval) + (1|ElephantName), data = soc.behav.rates)

summary(soc.lmer)

# residuals
plot(soc.lmer)
qqnorm(resid(soc.lmer))
qqline(resid(soc.lmer))

# residuals
residuals.soc.lmer <- simulateResiduals(fittedModel = soc.lmer)
plot(residuals.soc.lmer) # looks good!!!

testDispersion(residuals.soc.lmer) # is good
testOutliers(residuals.soc.lmer, type = "bootstrap") # looks good
testZeroInflation(residuals.soc.lmer) # looks good!

soc.lmer.emm <- ggpredict(soc.lmer, terms = c("SocialContext")) # to be used for creating Figure 2
```



## Vigilance
```{r}
vig.lmer <- lmerTest::lmer(LogBehavRate ~ SocialContext + (1|Year/EventInterval) + (1|ElephantName), data = vig.behav.rates)

summary(vig.lmer)

# residuals
plot(vig.lmer)
qqnorm(resid(vig.lmer))
qqline(resid(vig.lmer))

# residuals
residuals.vig.lmer <- simulateResiduals(fittedModel = vig.lmer)
plot(residuals.vig.lmer) # looks good!!!

testDispersion(residuals.vig.lmer) # is good
testOutliers(residuals.vig.lmer, type = "bootstrap") # looks good
testZeroInflation(residuals.vig.lmer) # looks good!

vig.lmer.emm <- ggpredict(vig.lmer, terms = c("SocialContext")) # to be used for creating Figure 2
```




# 3. Making Figure 2

## First have to adjust the ggpredict tables so they can all be merged
```{r}
# affiliation
affil.lmer.emm2 <- affil.lmer.emm %>%
  as.data.frame() %>%
  rename("SocialContext" = "x") %>%
  select(-group) %>%
  mutate(behavior = "affiliation")

# aggression
aggres.lmer.emm2 <- aggres.lmer.emm %>%
  as.data.frame() %>%
  rename("SocialContext" = "x") %>%
  select(-group) %>%
  mutate(behavior = "aggression")

# dominance
dom.lmer.emm2 <- dom.lmer.emm %>%
  as.data.frame() %>%
  rename("SocialContext" = "x") %>%
  select(-group) %>%
  mutate(behavior = "dominance")

# play
play.lmer.emm2 <- play.lmer.emm %>%
  as.data.frame() %>%
  rename("SocialContext" = "x") %>%
  select(-group) %>%
  mutate(behavior = "play")

# retreat
retreat.lmer.emm2 <- retreat.lmer.emm %>%
  as.data.frame() %>%
  rename("SocialContext" = "x") %>%
  select(-group) %>%
  mutate(behavior = "retreat")

# sda
sda.lmer.emm2 <- sda.lmer.emm %>%
  as.data.frame() %>%
  rename("SocialContext" = "x") %>%
  select(-group) %>%
  mutate(behavior = "sd.anxious")

# soc
soc.lmer.emm2 <- soc.lmer.emm %>%
  as.data.frame() %>%
  rename("SocialContext" = "x") %>%
  select(-group) %>%
  mutate(behavior = "social.content")

# vigilance
vig.lmer.emm2 <- vig.lmer.emm %>%
  as.data.frame() %>%
  rename("SocialContext" = "x") %>%
  select(-group) %>%
  mutate(behavior = "vigilance")

```


## Merge the tables
```{r}
lmer.emm <- full_join(affil.lmer.emm2, aggres.lmer.emm2) %>%
  full_join(., dom.lmer.emm2) %>%
  full_join(., play.lmer.emm2) %>%
  full_join(., retreat.lmer.emm2) %>%
  full_join(., sda.lmer.emm2) %>%
  full_join(., soc.lmer.emm2) %>%
  full_join(., vig.lmer.emm2)

# re-order the behaviors
lmer.emm2 <- lmer.emm[,c(6,1,2,3,4,5)]

lmer.emm2$behavior <- as.factor(as.character(lmer.emm2$behavior))
```


## Making the figure
```{r}
level_orded <- c("affiliation", "aggression", "dominance", "play", "retreat", "sd.anxious", "social.content", "vigilance")

behav.labs <- as_labeller(c("affiliation" = "Affiliation", "aggression" = "Aggression", "dominance" = "Dominance", "play" = "Play", "retreat" = "Retreat", "sd.anxious" = "Self-directed anxious", "social.content" = "Social contentment", "vigilance" = "Vigilance"))

# for adding in the significance stars - taken from Table A3 (output of models)
an.text <- data.frame(behavior = c("affiliation", "affiliation", "aggression", "dominance", "retreat", "sd.anxious", "social.content", "vigilance", "vigilance"), SocialContext = c("dom", "youth", "dom&youth", "youth", "musth&youth", "dom&youth","dom&youth", "dom&youth", "youth"), predicted = c(-4.2, -3, -3.85, -3.7, -3.8, -4.2, -4, -3.15, -2.85), label = c("*", "*", "*", "*", "*", "*", "*", "*", "*"))

ggplot(lmer.emm, aes(x = SocialContext, y = predicted, color = SocialContext)) +
  geom_count(show.legend = F) +
  facet_wrap(facets = "behavior", labeller = behav.labs, ncol = 4) +
  geom_pointrange(aes(x =SocialContext, y = predicted, ymin = conf.low, ymax = conf.high), linewidth = 1, size = 0.8) + 
  theme_bw(base_size = 19) +
  theme(legend.position = "none", 
        strip.background = element_rect(colour="black", fill="white", 
                                       size=1.5, linetype="solid")) +
  ylab(expression(Estimated~Mean~(Log[10])))+
  xlab(expression(Social~Context)) +
  scale_color_manual(labels = c("Adults", "Keystone", "Keystone & Youth", "Musth", "Musth & Youth", "Youth"), 
                    values = c("#9C755F", "#E15759", "#EDC948", "#59A14F", "#4E79A7", "#B07AA1")) +
  scale_x_discrete(labels = c("Adults", "Keystone", "Keystone & Youth", "Musth", "Musth & Youth", "Youth"), guide = guide_axis(angle = 45)) +
  geom_text(data = an.text, label = an.text$label, colour = "black", size = 10)

```


